---
title: "Assigning Weather Stations"
author: "Mike Logsdon"
date: "05/19/2015"
output: pdf_document
---


# The purpose of this document is to blow Ben Hannas' mind

## Access to Google Maps from Within rterm - Geocoding

The package rterm now includes access to the Google Maps Geo-coding API. This lets one 
quickly and with minimal oversight assign weather stations to a large number of buildings
by address. There will inevitably be some ambiguities and errors with this 
hands-off approach, but at the very least should be a starting point to greatly speed up 
the task of associating buildings with weather files.


The code to geocode is actually pretty simple. Here is the rterm function. Notice the 
Sys.sleep command at the end. Google Maps only allows 10 queries per second (at least
with free access), so it gets manually slowed down so that you don't unexpectedly retrieve 
a bunch of NA values by maxing out your allotted queries. Also note that, similarly to the 
NOAA read.ghcn stuff, it assumes that you have registered and activated a Google Maps Geocoding
API key, and that it lives in your global environment under the name "google_key". 
[This page](https://developers.google.com/maps/documentation/geocoding/?hl=en_US#api_key) will walk you through the process of registering a Google Maps API key and activating the Geocoding API.

```{r, eval=FALSE}
getGeoCode <- function(name, quietly = TRUE) {
  if(!exists("google_key")) {
    stop("Must Register a Google Maps API Key to Search by City/Address")
  }
  name <- enc2utf8(gsub(" ", "%20", name))
  urlx <- paste("https://maps.google.com/maps/api/geocode/json?sensor=false&address=", 
                      name, "&key=", google_key, sep = "")

  if(!quietly) print(paste("Attempting to query:", urlx))
  tmp <- httr::GET(urlx)
  results <- jsonlite::fromJSON(httr::content(tmp, "text"))$results
  
  geocode <- as.numeric(results$geometry$location)
  names(geocode) <- c("lat", "lon")

  Sys.sleep(.15)
  return(geocode)
}
```



### Access to Google Maps from Within rterm - creating maps

We have also plugged rterm into the [R ggmap package](http://cran.r-project.org/web/packages/ggmap/index.html) from David Kahle and Hadley Wickham. 


```{r}
library(rterm)
stations <- stationSearch("4056 9th Ave NE Seattle WA")
stations
```

```{r, fig.width=11, fig.height=9}
stationMap(stations)
```

```{r, fig.width = 10, fig.height = 8}
comp <- stationCompare(stations, "2013-01-01", "2013-12-31")
stationMap(comp)
```



### An example of bulk Assignments


Here's an example with the DHP data of the function "assignStations", which is blindly looking up the closest NOAA GHCN Daily weather station to the specified city name.

```{r}
library(foreign)
library(plyr)
library(dplyr)


setwd("/storage/server/NEEA/DHP_Impact_Evaluation/4_Billing_Analysis/Final_Dataset/data_and_analysis/code/")

# Read and process the data
dset <- read.dta(file = "../inputs/Full_Dataset.dta") %>%
  mutate(kwhd = as.numeric(kwhd)) %>%
  mutate(avetemp = as.numeric(avetemp)) %>%
  mutate(kwh = as.numeric(kwh)) %>%
  mutate(readdate = as.Date(readdate, format = "%m/%d/%Y")) %>%
  mutate(idPrePost = interaction(id, post)) %>%
  filter(!is.na(post))
  

# Create map of site id to location
locations <- ddply(dset, .(id), function(x) {
  c("location" = paste(x$city[1], x$state[1], sep = ", "))
})

# As a demonstration
loc10 <- locations[1:10, ]
loc10
assignStations(loc10, "location")
```

### Follow-up on an oddity
